Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-09-02 paper/main.tex
formulate new introduction

Despite their popularity, \acp{abm} face multiple challenges in practise.
Constructing a fully working \ac{abm} simulation from scratch is an intricate process that requires
a large set of skills and time.
It is therefore common practise to reuse existing simulation tools, which brings along a number of
other problems.
Even if the existing tool is designed to be generalizable, it often comes with a pre-defined set
of parameters that need to be specified.
Furthermore, as we have discussed in our previous work~\cite{Pleyer2023}, these tools were most
often not designed to be flexible but rather for specific purposes thus limiting their
generalizability.
% TODO citations
Typically researchers aim to estimate every individual parameter of their model.
This is achieved using methods such as the profile-likelihood and model reduction techniques.
% TODO citaitons
When it comes to \acp{abm} however, it is often only shown that the results which have been
generated by an \ac{abm} lead to the phenomenon which aligns with the observed biological reality.
% TODO citations
This reduces the confidence in the constructed models and raises concerns with respect to the
validity of the obtained results.

>>>>

Despite their popularity, \acp{abm} face multiple challenges in practise.
Constructing a fully working \ac{abm} simulation from scratch is an intricate process that requires
a large set of skills and time.
It is therefore common practise to reuse existing simulation tools, which brings along a number of
other problems.
Even if the existing tool is designed to be generalizable, it often comes with a pre-defined set
of parameters that need to be specified.
Furthermore, as we have discussed in our previous work~\cite{Pleyer2023}, these tools were most
often not designed to be flexible but rather built for specific purposes thus limiting their
generalizability.
% TODO citations
Typically researchers aim to estimate every individual parameter of their model.
This is achieved using methods such as the profile-likelihood and model reduction techniques.
% TODO citaitons
When it comes to \acp{abm} however, it is often only shown that the results which have been
generated by an \ac{abm} lead to the desired phenomenon which aligns with the observed biological
reality thus omitting to actually estimate the necessary parameters directly.
% TODO citations
This reduces the confidence in the constructed models and raises concerns with respect to the
validity of the obtained results.

In this work, we address these challenges using a general-purpose mechanical model for rod-shaped
bacteria.
Our approach provides various methods which can be combined to systematically estimate the
parameters of our model.
We deliberately distinguish between conceptually separate components within this framework.
This modularity highlights a "blueprint" which can be adapted and applied to other \acp{abm}.

Mechanical properties of rod-shaped bacteria have been studied
extensively~\cite{Chatterjee1988,Takeuchi2005,IWAI2002}.
It was found that the MreB protein is the driving force in maintaining their distinct elongated
shape~\cite{Ursell2014}, which plays an important role in the dynamics of
growth~\cite{Billaudeau2017}, division~\cite{Harry2001} and their collective
behaviour~\cite{vanGestel2015}.
Furthermore, it has been discovered that the rod can be bent thus introducing temporary elastic or
permanent plastic deformations~\cite{Amir2014_2}.
They are an interesting target system since their physical form provides more surface information
compared to spherical cells.
In addition, parameters such as growth rates have been found to vary across multiple
cells~\cite{Koutsoumanis2013} which highlights the need to incorporate this heterogeneity on a
single-cell level.

Computational models for rod-shaped bacteria allow researchers to systematically analyze their
properties and infer emerging phenomena~\cite{Cho2007,Jnsson2005}.
Existing tools can describe growth, division and many effects such as colony formation and
expansion~\cite{Matyjaszkiewicz2017,Doumic2020} but they are limited in describing the
aforementioned plastic or elastic deformations of the rod.
Furthermore, the literature on how to estimate the parameters of such models is very sparse.

This work adresses some of these limitations.
We use microscopic image time series as inputs to obtain information about the spatial
configuration of the cells.
This information can be used as initial values for a numerical simulation.
By varying the parameters of the model and comparing the results to the microscopic data, we can
estimate and quantify the parameters of our model.
Our approach takes a general stance which incorporates effects that are observed within
\textit{B.Subtilis} and \textit{E.Coli}.
